---
title: "Exercise5"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Complex Contagion
- How do we explain madness/wisdom of crowds?
- Focusing on individuals inherently gets us 'stuck in the weeds' when trying to explain macro-scale trends
- Network analysis allows us to study the connections of people
instead of the people themselves

- This helps to explain certain paradoxes (Majority Illusion), like for eg when everyone in a given social network believes that everyone else has more friends than them. This seems possible but is just a consequence of people only being able to see their own part of the network, not the whole

- The concept of Majority Illusion could also explain things like group think (eg everyone around me has the same political leanings)

- How do we define 'Contagion'? In this instance, it isn't necessarily all negative things like misinformation or madness, it can also be healthy lifestyles, safety information, etc. Contagion is just something that can be quickly spread across a connected network.







```{r}


```




