---
title: "Exercise3"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Libraries
```{r}
library(ggraph)
library(igraph)

library(arrow)
library(tidyverse)
library(gender)
library(wru)
library(lubridate)

library(ggplot2)
library(gridExtra)
library(grid)
```

## Data
```{r}
# change to your own path!
data_path <- "Data/"
applications <- read_parquet(paste0(data_path,"app_data_sample.parquet"))
edges <- read_csv(paste0(data_path,"edges_sample.csv"))

```

## Add gender
```{r}
# get a list of first names without repetitions
examiner_names <- applications %>% 
  distinct(examiner_name_first)

# get a table of names and gender
examiner_names_gender <- examiner_names %>% 
  do(results = gender(.$examiner_name_first, method = "ssa")) %>% 
  unnest(cols = c(results), keep_empty = TRUE) %>% 
  select(
    examiner_name_first = name,
    gender,
    proportion_female
  )
# remove extra columns from the gender table
examiner_names_gender <- examiner_names_gender %>% 
  select(examiner_name_first, gender)

# joining gender back to the dataset
applications <- applications %>% 
  left_join(examiner_names_gender, by = "examiner_name_first")

# cleaning up
rm(examiner_names)
rm(examiner_names_gender)
gc()

```

## Add race
```{r}
# get list of distinct last names
examiner_surnames <- applications %>% 
  select(surname = examiner_name_last) %>% 
  distinct()

examiner_race <- predict_race(voter.file = examiner_surnames, surname.only = T) %>% 
  as_tibble()

# infer racial probabilities from surname tibble
examiner_race <- examiner_race %>% 
  mutate(max_race_p = pmax(pred.asi, pred.bla, pred.his, pred.oth, pred.whi)) %>% 
  mutate(race = case_when(
    max_race_p == pred.asi ~ "Asian",
    max_race_p == pred.bla ~ "black",
    max_race_p == pred.his ~ "Hispanic",
    max_race_p == pred.oth ~ "other",
    max_race_p == pred.whi ~ "white",
    TRUE ~ NA_character_
  ))

# removing extra columns and merge into application data
examiner_race <- examiner_race %>% 
  select(surname,race)

applications <- applications %>% 
  left_join(examiner_race, by = c("examiner_name_last" = "surname"))
# cleanup
rm(examiner_race)
rm(examiner_surnames)
gc()
```

## Add tenure
```{r}
# get all application filing dates
examiner_dates <- applications %>% 
  select(examiner_id, filing_date, appl_status_date) 

# calculate start and end date from filing / status date respectively
examiner_dates <- examiner_dates %>% 
  mutate(start_date = ymd(filing_date), end_date = as_date(dmy_hms(appl_status_date)))

# for each examiner, get earliest and latest days, then interval between them as tenure in days
examiner_dates <- examiner_dates %>% 
  group_by(examiner_id) %>% 
  summarise(
    earliest_date = min(start_date, na.rm = TRUE), 
    latest_date = max(end_date, na.rm = TRUE),
    tenure_days = interval(earliest_date, latest_date) %/% days(1)
    ) %>% 
  filter(year(latest_date)<2018)

# merge and clean
applications <- applications %>% 
  left_join(examiner_dates, by = "examiner_id")

rm(examiner_dates)
gc()
```

## Subset two workgroups for analysis

```{r}

wg163 = applications[substr(applications$examiner_art_unit, 1,3)==163,]
wg176 = applications[substr(applications$examiner_art_unit, 1,3)==176,]

wg163

```
## Demographics analysis

```{r}
# summary statistics

summary(wg163)
summary(wg176)

```
Tenure range: 339-6350

```{r}

a <- ggplot(data=wg163, aes(x=gender)) +
  geom_bar(aes(y = (..count..)/sum(..count..)) )  +
  ylab("Proportion")+
  xlab("Gender")+
  ylim(0,1)+
  ggtitle(paste0("Gender distribution for wg 163"))

b <- ggplot(data=wg176, aes(x=gender)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  ylab("Proportion")+
  xlab("Gender")+
  ylim(0,1)+
  ggtitle(paste0("Gender distribution for wg 176"))
grid.arrange(a,b,ncol=2, widths=c(1,1))


```
We notice 176 has a much more male-skewed population, lets now repeat for race...


```{r}
a <- ggplot(data=wg163, aes(x=race)) +
  geom_bar(aes(y = (..count..)/sum(..count..)) )  +
  ylab("Proportion")+
  xlab("Race")+
  ylim(0,1)+
  ggtitle(paste0("Race distribution for wg 163"))

b <- ggplot(data=wg176, aes(x=race)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  ylab("Proportion")+
  xlab("Race")+
  ylim(0,1)+
  ggtitle(paste0("Race distribution for wg 176"))
grid.arrange(a,b,ncol=2, widths=c(1,1))


```
Both are predominantly white, though perhaps 176 by a bit smaller margin.

We could choose to look at both race and gender at the same time:
```{r}
a <- ggplot(data=wg163, aes(x=race)) +
  geom_bar(aes(y = (..count..)/sum(..count..), fill=gender  ), show.legend=FALSE )  +
  ylab("Proportion")+
  ylim(0,1)+
  xlab("Race")
b <- ggplot(data=wg176, aes(x=race)) +
  geom_bar(aes(y = (..count..)/sum(..count..), fill=gender  ) )  +
  ylab("Proportion")+
  ylim(0,1)+
  xlab("Race")
grid.arrange(a,b,ncol=2, widths=c(1,1.4), top=textGrob("Race distribution for work groups 163 & 176 colored by examiner gender"))

```
In this view, the male-female unbalance among 176-ers is mostly accounted for by the white population. 


## Advice networks

```{r}
# first get work group for each examiner and limit to our two wgs of interest
examiner_aus = distinct(subset(applications, select=c(examiner_art_unit, examiner_id)))
# we eventually want to make a network with nodes colored by work group, so lets add that indicator
examiner_aus$wg = substr(examiner_aus$examiner_art_unit, 1,3)
# restrict down to our selected art units to reduce merging complexity later on
examiner_aus = examiner_aus[examiner_aus$wg==163 | examiner_aus$wg==176,]

# now we will merge in the aus df on applications 
tM = merge(x=edges, y=examiner_aus, by.x="ego_examiner_id", by.y="examiner_id", all.x=TRUE)
tM = tM %>% rename(ego_art_unit=examiner_art_unit, ego_wg=wg)
# drop the examiner edges who are not in 163 or 176
tM = drop_na(tM)

# now repeat for the alter examiners
tM = merge(x=tM, y=examiner_aus, by.x="alter_examiner_id", by.y="examiner_id", all.x=TRUE)
tM = tM %>% rename(alter_art_unit=examiner_art_unit, alter_wg=wg)
tM = drop_na(tM)
# we are left with 870 edges corresponding to instances of examiners in wg163 or wg176 asking for advice


egoNodes = subset(tM, select=c(ego_examiner_id,ego_art_unit, ego_wg)) %>% rename(examiner_id=ego_examiner_id,art_unit=ego_art_unit,wg=ego_wg)
alterNodes = subset(tM, select=c(alter_examiner_id,alter_art_unit, alter_wg))%>% rename(examiner_id=alter_examiner_id,art_unit=alter_art_unit,wg=alter_wg)
nodes = rbind(egoNodes, alterNodes)
nodes = distinct(nodes)

# problem: when we reduce to the list of distinct vertices, we actually have more than we should, since some examiners move amongst art units/wgs in this data subset
nodes = nodes %>% group_by(examiner_id) %>% summarise(examiner_id=first(examiner_id), art_unit=first(art_unit), wg=first(wg))

```


```{r}
adviceNet = graph_from_data_frame(d=tM, vertices=nodes, directed=TRUE)
adviceNet

```
```{r}
# Calculate centralities and add as vertex attributes
Degree <- degree(adviceNet, v=V(adviceNet))
Betweenness <- betweenness(adviceNet)
Eigenvector <- evcent(adviceNet)$vector

V(adviceNet)$size = Degree
V(adviceNet)$eig = round(Eigenvector,2)
V(adviceNet)$bet = round(Betweenness,2)

V(adviceNet)$color = nodes$art_unit
```

```{r}
#E(adviceNet)

```


```{r}
centralities <- cbind(Degree, Eigenvector, Betweenness)
centralities = round(centralities,2)
centralities = data.frame(centralities)
```

```{r}
V(adviceNet)
```

```{r}

ggraph(adviceNet, layout="kk") +
  geom_edge_link()+
  geom_node_point(aes(size=size, color=color), show.legend=T)
  #geom_node_text(aes(label=name))

```
Interestingly we do seem to have several instances of an examiner leaving the confines of their workgroup and seeking advice from the other (see the small components on the outer edges with both blue and black nodes). 

When we explore the centralities data frame, we also see there are two nodes with much higher degree than the rest, examiner 72253 and examiner 84157. Who are these people?

```{r}
applications[applications$examiner_id==72253,]

```
So we know this is the node at the centre of the bottom network component (wg163 for this examiner). Recall from our summary statistics earlier that examiner tenure ranged from roughly 340 to 6350. Note examiner 72253 has a tenure of 6349(!) this provides a logical explanation for why this node appears to be so important (at least by degree): they've been around the longest and have had many more opportunities to seek help. Note also that this examiner has the highest betweenness score in the network, implying they are the important for connecting disparate parts of the advice network.


```{r}
applications[applications$examiner_id==84157,]

```
Examiner 84157 actually did not start their career with wg 176, but rather 171. They must have moved at some point during their tenure. Nevertheless, we can be quite sure that this is the large node in the top blue component representing advice for wg176. As expected, the tenure for this examiner is quite long as well - 6339 days, so this examiner has also had the longest possible amount of time to ask for advice during their career. 

A dichotomy arises between these two examiners of interest - it appears that 84157 has the highest eigenvector centrality, while 72253 has the highest betweenness centrality. I would posit this could be due to wg176 having a much larger advice network overall, with several long-tenured members. This would increase the eigenvector centrality for 84157 (because it has important friends) while simultaneously decreasing their betweeness centrality (since there are many paths across the wg176 network)

The opposite is true for examiner 72253: Their friends are not particularly "important" so they receive a low eigenvector centrality, but since the network is fairly large and somewhat "star" shaped, theya re very important for network traversals.


