---
title: "Exercise1"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Imports
```{r}
# graph analysis
library(tidygraph)
library(ggraph)
library(igraph)

# tidy data analysis and visualziation
library(readr)
library(dplyr)
library(skimr)
```


## Read Data
```{r}
connections = read.csv("Data\\Connections.csv")
```


## Cleaning
```{r}
connections = subset(connections, select=-c(Email.Address))
connections %>% skim() #good to go
```

## Preprocessing
```{r}
connections$Nodes = paste(connections$First.Name, substr(connections$Last.Name,0,1)) # Node names are first + first letter of last name


```



## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
